Supplementary Information

### 2016-12-05 Response_to_reviewers.Rmd

> Lines 231-232: Looking at the maps, it seems that this statement is limited by the fact that the current study does not incorporate the northern range limit for many species. Can the authors address this limitation at some point?

We have added the paragraph:  *For all species, the range within the upper Midwest incompletely samples species ranges, however, the region contains several major ecotones and the southern and western range limit for many tree species [@curtis1959vegetation] so this region comprises an important boundary in geographic and climate space.* to address this point.

> Lines 337-351: The authors discuss agriculture, but not timber harvesting and changes in fire regime in terms of the land-use drivers. It seems that the latter two are potentially important for some of the regional patterns. Additionally, I wonder if some of this could be discussed by looking at gains and losses relative to NLCD land-use layers. Perhaps that was in the supplement and I missed it.

We have added the sentence: *While logging and altered fire regimes have altered community composition in the north, they have not widely excluded taxa from the northern regions and associated climates, except for * Tsuga * (Figure 3), a shade-tolerant and late-successional taxon.*

@@

> Figure 5: The x – axis needs more tick marks and numbers. At first, my mind assumed a linear scale, which would imply negative distances, but then I thought maybe this was some sort of log scale, which would make more sense with a distance metric. Some additional clarity is needed.

We have changed the $x$-axis to a $log_{10}$ scale and added more tick marks to the $x$-axis to make it clear that it is non-linear.

@@

> 183-185   First, it is a serious stretch to use 1895-1919 climate to represent tree/climate relationships during 1830-1910.  The verbal statement (without statistical analysis) that the dendro-based PDSI did not change over that period is not very convincing.  Moreover, the analysis specifically should be done on the climate predictors used in the niche modeling.  Also, evaluate to what extent does this PDSI index represent trends in Tmin and the other predictors used?

We respectfully disagree that it is a serious stretch to use 1895-1915 climates to represent 1830-1910 climates.  The PDSI data shows no evidence of long-term trend between 1800 and 1910, based on a LOESS line fitted to the time series shown in Fig. 1 (see other responses below).

To further address this point, we have added a review of early meteorological records from US military forts. These records do indicate some warming in the region (Baker et al, 1985, Burnett et al. 2010).   We have added the following text to the first paragraph of the results.

*Early temperature observations from US military forts indicate warming between the middle 19^th^ and early 20^th^ centuries at Fort Snelling, Minnesota (Baker et al. 1985) and Fort Leavenworth, Kansas (Burnette et al. 2010). The latter shows faster increases during winter (0.10^o^C decade-1) than summer (0.07^o^C decade^-1^). Hence, our estimates of historic climate change may be too conservative for $T_{min}$ and $T_{diff}$ and too large for $T_{max}$ (Supplementary Information).*

@@

See responses to prior comments.

> 4) The 2×2 factorial design with pre-settlement and modern vegetation and climates (VHCH, VHCM,VMCH,VMCM) was very confusing (Figure 2a). The historical distribution data for VHCH scenario were derived from PLS and the contemporary distribution data for VMCM were derived from US FIA data. Then, what were the data for VHCM and VMCH scenarios. It was impossible to get the data for VHCM and VMCH scenarios, right? If my understanding was wrong, then the more detailed information was needed to better describe the factorial design and data used for each scenario.

We revised this section of the methods for clarity, in part by re-writing this paragraph:

*The total observed shift (d_tot) between a historical and modern PDF is the Hellinger distance between V_M C_M and V_H C_H (Figure 2a: top left to bottom right). The conditions V_M C_H and V_H C_M are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and *vice versa*. The effect of climate change on shifting PDFs (d_c) is the difference between PDFs using the early climate normals (C_H) and modern climate normals (C_M) overlaid on PLS vegetation (e.g., V_H C_H versus V_H C_M; Figure 2a; left column). The effect of vegetation change on shifting PDFs is d_v, the Hellinger distance between V_H C_H and V_M C_H (top row of Figure 2a). Because Euro-American land-use is the dominant driver of vegetation change in the upper Midwest over the past two centuries (Introduction), we often attribute shifts recorded by d_v to the land-use change associated with Euro-American settlement.*

@@

> 5) About the t-tests to assess the interaction effects between land use and climate change (e.g., compounding, confounding, counteracting), what the responsible variables were used for the comparisons between the VHCH and VHCM, VHCH and VMCH. More detailed information was needed to better describe these tests.

We have modified the t-tests to include spatial autocorrelation and expanded their description in the Supplementary Information.  The responsible variables are explained more clearly in the methods now.  We compare VHCH and VHCM to estimate the significance and direction of the climate shift (vegetation is fixed, climate layer changes), and compare VHCH to VMCH to estimate the change in vegetation (keeping vegetation constant).

@@

> 4) The figures were not well made. The axis titles in Figure 1 were two small. The four lines in the Figure 2 b-e were difficult to read. The genera names in Figure 5 were too crowd and hard to tell. It may be better to use shorter names instead.

We have edited the figures to address these comments.

> 5) All figure titles were too long. Many of these contents were approach and should be put in methods section.

We have added more information to the Methods section, while also retaining much of this information in the Figure legends.  We prefer a style that uses longer legends, to facilitate figure interpretation.
